Implementation of a minimum branch-decomposition and branch-width algorithm for planar graphs.

### 2024-09-02 paper/main.tex
Add graphs

\begin{center}
	\includegraphics[width=0.9\textwidth]{images/dalle-pied-piper.png}
\end{center}
\footnote{Illustration by Dall-E via ChatGPT (GPT-4o) with the prompt "Generate an image of pied piper the rat-catcher along with a computer science graph".}

@@

	With subroutines for all three problems, the implementation obtains a branch decomposition like so.

	\pyfile{branch_decomposition.py}{19}{27}

@@

		\pyfile{medial_graph.py}{4}{54}

@@

			\pyfile{carving_width.py}{5}{6}
			\pyfile{carving_width.py}{130}{156}

@@

		The rat-catching algorithm decides $cw(M) \geq k$ with $k \in \mathbb{N}^+$.

		The rat-catching algorithm can be described as a game of two players, the rat and rat-catcher. Considering a graph $M$, the edges of a face can be thought of as walls of a room and vertices as the corners of some rooms. The rat moves from corner to corner along the walls and the rat-catcher moves from room to room through some wall. The rat-catcher can force the rat away from some walls by making noise. A round of this game is played with some noise level $k$. The rat-catcher wins the round if they can force the rat to be in some wall of the room that they are in, and the rat wins the round if there is a strategy whereby the rat can escape indefinitely.

		Additionally, if $\Delta(M) \geq k$ then the rat wins. The argument for why this is true is glossed over in \ref{BGZ15}.

@@

		A \textit{quiet subgraph} $Q(M, k, e)$, for some graph $M$ and some noise level $k$ and some edge $e \in \mathbb{E}(M)$, is a subgraph of $M$ with the vertex set $V(Q(M, k, e)) = V(M)$ and the edge set

		$\mathbb{E}(Q(M, k, e)) = \{e_1 \st $ every closed walk of $M^*$ containing $e_1^*$ and $e_2^*$ has length at least $k\}$

		\begin{problem}\label{problem:noisy-subgraph}
			Given a plane graph $M$ that might have parallel edges, an edge $e \in \mathbb{E}(M)$, and noise level $k \in \mathbb{N}^+$, output the quiet subgraph $Q(M, k, e)$.
		\end{problem}

@@

		Let $F(M)$ be the set of faces of $M$.

@@

		Let $T$ be every possible state when the rat-catcher is on an edge. $T = \{ (e, C) \st e \in \mathbb{E}(M) \land C$ is a component of $Q(M, k, e) \}$.

@@

		You can mark vertices of the graph $H$ that are losing states, and then repeatedly mark any state that leads to a losing state, until either every state is marked or no more states can be marked. If every state is marked then the rat-catcher wins, otherwise the rat wins.

@@

		Losing states (the tuples $(f,v) \in S$ where $v \in f$) are marked as losing.

@@

		\pyfile{carving_width.py}{79}{128}

@@

			Using the mapping from links to edges, and the fact that an edge $e$ is called quiet iff. $e$ is not noisy, the quiet edges can be obtained by applying the map.

@@

			\pyfile{carving_width.py}{6}{74}

@@

@@

	The first set of graphs are random cubic planar graphs. The second set of graphs are small simple planar graphs generated by \textit{Plantri}\footnote{\url{http://users.cecs.anu.edu.au/~bdm/plantri/}}\footnote{\url{https://github.com/mishun/plantri/tree/master}}.

@@

	Some of the graphs are suspected to be cause the implementation to go into an infinite loop, as the branch width is not computed within a reasonable time. So the implementation is suspectedly not correct for all planar graphs.

	\begin{figure}[H]
		\centering
		\includegraphics[width=\textwidth]{./images/bw-planar.png}
		\caption{Branch width of planar graphs generated by Plantri. The x-axis is the number of vertices in the graph, and the y-axis is the branch width. The red line is $\lfloor 4.5 \cdot \sqrt{n} \rfloor$. The blue dots are the branch width and the number directly next to it is the number of the graph with the same unique (n, bw) pair.}
		\label{fig:bw-planar}
	\end{figure}

@@

\newpage
\section{References}
	\printbibliography

\newpage
\section{Appendix}
	\subsection{}
	\pyfile{branch_decomposition.py}{0}{1000}\label{branch_decomposition.py}
	\subsection{}
	\pyfile{branch_width.py}{0}{1000}\label{branch_width.py}
	\subsection{}
	\pyfile{carving_decomposition.py}{0}{1000}\label{carving_decomposition.py}
	\subsection{}
	\pyfile{carving_width.py}{0}{1000}\label{carving_width.py}
	\subsection{}
	\pyfile{contraction.py}{0}{1000}\label{contraction.py}
	\subsection{}
	\pyfile{dual_graph.py}{0}{1000}\label{dual_graph.py}
	\subsection{}
	\pyfile{Graph.py}{0}{1000}\label{Graph.py}
	\subsection{}
	\pyfile{medial_graph.py}{0}{1000}\label{medial_graph.py}
	\subsection{}
	\pyfile{test-cubic-cross-validate.py}{0}{1000}\label{test-cubic-cross-validate.py}
	\subsection{}
	\pyfile{test-planar-graphs.py}{0}{1000}\label{test-planar-graphs.py}
	\subsection{}
	\pyfile{util.py}{0}{1000}\label{util.py}
	\subsection{}
	\pyfile{graphs/gen-random-planar-cubic.py}{0}{1000}\label{graphs/gen-random-planar-cubic.py}
	\subsection{}
	\pyfile{graphs/gen-planar.py}{0}{1000}\label{graphs/gen-planar.py}
	\subsection{}
	\pyfile{graphs/process_planar_code.py}{0}{1000}\label{graphs/process_planar_code.py}
	\subsection{}
\end{document}

>>>>

\begin{center}
	\includegraphics[width=0.9\textwidth]{images/dalle-pied-piper-2.png}
\end{center}
\footnote{Illustration by Dall-E via ChatGPT (GPT-4o) with the prompt "Generate an image of pied piper the rat-catcher along with a computer science graph".}

@@

	With subroutines for all three problems, the implementation obtains a branch decomposition.

	\pyfile{branch_decomposition.py}{20}{25}

@@

		\pyfile{medial_graph.py}{4}{50}

@@

			\pyfile{carving_width.py}{5}{5}
			\pyfile{carving_width.py}{130}{156}

@@

		The rat-catching algorithm decides $cw(M) \geq k$ with $M$ being a graph and $k \in \mathbb{N}^+$.

		The rat-catching algorithm can be described as a game of two players, the rat and rat-catcher. Considering a graph $M$, the edges of a face can be thought of as walls of a room and vertices as the corners of some rooms. The rat moves from corner to corner along the walls and the rat-catcher moves from room to wall to room again. The rat-catcher can force the rat away from some walls by making noise. A round of this game is played with some noise level $k$. The rat-catcher wins the round if they can force the rat to be in some wall of the room that they are in, and the rat wins the round if there is a strategy whereby the rat can escape indefinitely.

		Additionally, if $\Delta(M) \geq k$ then the rat wins. The argument for why this is true, is glossed over in \cite{BGZ15}, but is important to include.

@@

		A \textit{quiet subgraph} $Q$, for some graph $M$ and some noise level $k$ and some edge $e \in \mathbb{E}(M)$, is a subgraph of $M$ with the vertex set $V(Q) = V(M)$ and the edge set

		$\mathbb{E}(Q) = \{e_1 \st $ every closed walk of $M^*$ containing $e_1^*$ and $e_2^*$ has length at least $k\}$

		\begin{problem}\label{problem:noisy-subgraph}
			Given a plane graph $M$ that might have parallel edges, an edge $e \in \mathbb{E}(M)$, and noise level $k \in \mathbb{N}^+$, output the quiet subgraph $Q$.
		\end{problem}

@@

@@

		Let $T$ be every possible state when the rat-catcher is on an edge. $T = \{ (e, C) \st e \in \mathbb{E}(M) \land C$ is a component of $Q \}$.

@@

		You can mark vertices of the graph $H$ that are losing states, and then repeatedly mark any state that leads to a losing state, until either every state is marked or no more states can be marked. If every state is marked then the rat-catcher wins, otherwise the rat wins. Exactly how to propagate losing states is discussed below.

@@

		The initial losing states (the tuples $(f,v) \in S$ where $v \in f$) are marked as losing.

@@

		Apply the two rules above until either no more new losing states are marked, or until every state is marked as losing.

		It is useful for the implementation to have a relation \mintinline{py}{edge_to_node} mapping the two half-edges of an edge to the two faces to either side of the edge, and a relation \mintinline{py}{node_to_face} to enumerate the faces. These relation are computed in the \mintinline{py}{dual_graph} function.

		\pyfile{carving_width.py}{5}{6}
		\pyfile{carving_width.py}{79}{128}

@@

			Using the relation \mintinline{py}{link_to_edge} and the fact that an edge $e$ is called quiet iff. $e$ is not noisy, the quiet edges can be obtained by applying the relation.

@@

			\pyfile{carving_width.py}{5}{74}

@@

			The relations \mintinline{py}{node_to_face} and \mintinline{py}{edge_to_node} are useful for previously mentioned subproblems, and are computed here while enumerating the faces.

@@

	The first set of graphs are random cubic planar graphs. The second set of graphs are small simple planar graphs generated by \textit{Plantri}\footnote{\url{http://users.cecs.anu.edu.au/~bdm/plantri/}}\footnote{\url{https://github.com/mishun/plantri/tree/master}}.\cite{BM}

@@

	Some of the graphs are suspected to cause the implementation to go into an infinite loop, as the branch width is not computed within a reasonable time. So the implementation is suspectedly not correct for all planar graphs.

	\begin{figure}[H]
		\centering
		\includegraphics[width=\textwidth]{./images/bw-planar.png}
		\caption{Branch width of planar graphs generated by Plantri. The x-axis is the number of vertices in the graph, and the y-axis is the branch width. The red line is $\lfloor 4.5 \cdot \sqrt{n} \rfloor$. The blue dots are the branch width and the number directly next to it is the number of the graph with both the same number of vertices and the same branch width.}
		\label{fig:bw-planar}
	\end{figure}

@@

\section{Acknowledgements}
	I would like to thank my supervisors, Thore Husfeldt and Andreas Björklund, for their guidance and support throughout this project. I would also like to thank Andreas Björklund for providing the random cubic planar graphs and values for cross-validation. I would like to thank my family and friends for their support and encouragement.

\newpage
\printbibliography

\newpage
\section{Appendix}
	\subsection{}
	\pyfile{branch_decomposition.py}{0}{1000}\label{branch_decomposition.py}
	\subsection{}
	\pyfile{branch_width.py}{0}{1000}\label{branch_width.py}
	\subsection{}
	\pyfile{carving_decomposition.py}{0}{1000}\label{carving_decomposition.py}
	\subsection{}
	\pyfile{carving_width.py}{0}{1000}\label{carving_width.py}
	\subsection{}
	\pyfile{contraction.py}{0}{1000}\label{contraction.py}
	\subsection{}
	\pyfile{dual_graph.py}{0}{1000}\label{dual_graph.py}
	\subsection{}
	\pyfile{Graph.py}{0}{1000}\label{Graph.py}
	\subsection{}
	\pyfile{medial_graph.py}{0}{1000}\label{medial_graph.py}
	\subsection{}
	\pyfile{test-cubic-cross-validate.py}{0}{1000}\label{test-cubic-cross-validate.py}
	\subsection{}
	\pyfile{test-planar-graphs.py}{0}{1000}\label{test-planar-graphs.py}
	\subsection{}
	\pyfile{util.py}{0}{1000}\label{util.py}
	\subsection{}
	\pyfile{graphs/gen-random-planar-cubic.py}{0}{1000}\label{graphs/gen-random-planar-cubic.py}
	\subsection{}
	\pyfile{graphs/gen-planar.py}{0}{1000}\label{graphs/gen-planar.py}
	\subsection{}
	\pyfile{graphs/process_planar_code.py}{0}{1000}\label{graphs/process_planar_code.py}
\end{document}
